---
title: "TriLLIEM"
subtitle: "Triad Log-Linear modelling of Imprinting, Environmental interactions, and Maternal effects"
author: Kevin Zhao
output:
  beamer_presentation:
    theme: "AnnArbor"
    colortheme: "dolphin"
    fonttheme: "structurebold"    
header-includes: 
  - \usepackage{tikz}
  - \usepackage{pgfplots}
  - \usepackage{xcolor}
  - \pgfplotsset{compat=1.18}
  - \usepackage{mathptmx}
  - \usepackage[usenames,dvipsnames]{xcolor}
  - \usepackage{booktabs}
  - \let\oldverbatim\verbatim
  - \let\endoldverbatim\endverbatim
  - \renewenvironment{verbatim}{\footnotesize\oldverbatim}{\endoldverbatim}
classoption:
  - twocolumn
---

```{r setup, include = FALSE}
library(tidyverse); theme_set(theme_bw())
library(TriLLIEM)
library(parallel)
library(knitr); options(knitr.kable.NA = '')
library(kableExtra)
```

## Triad data

- Goal: Quantifying the effect of a single-nucleotide polymorphism (SNP) of interest on one's disease risk.
- Case-control studies (\color{blue}Control \color{black}+ \color{red}Case\color{black}) are decent, but influenced by \textbf{population substructures}.
- Case-parent/Triad studies (\color{blue}Control Parents \color{black}+ \color{red}Case Child\color{black}) address this, and can identify \emph{maternal effects}.  Can't quantify mating asymmetry with case-triads alone.
- A \textbf{hybrid} triad study with \color{blue}control \color{black}and \color{red}case \color{black}children can solve this issue.

## The log-linear model

- The counts of each triad are modelled with a log-linear GLM.
- A simple example of a model we can apply to data:
\[\ln(n_{M,F,C,D}) = \ln (\mu_i) + CD \ln R + MD \ln S + \ln (\mathrm{Off}).\]
- Additional terms to consider:
    - Mating type parameters (for HWE, MS, and MaS)
    - Imprinting effects ($I_{\rm M}$ and/or $I_{\rm F}$)
    - \textbf{Gene-environment interactions} ($\textrm{M}:\textrm{E}$, $I_{\rm M}:\textrm{E}$, $I_{\rm F}:\textrm{E}$)

## Population stratification

- Population stratification occurs when there are distinct sub-populations in the population we're sampling from.
- Sub-populations ($X = i$) could respond to the SNP of interest the same, but have different demographic parameters, such as:
    - Prevalence of the minor allele (${\rm MAF}_i$),
    - Prevalence of the disease ($\Pr(D = 1 | X = i) = \alpha_i$),
    - Rates of environmental exposure for both cases and controls ($\Pr(E = 1 | X = i, D = 1) = \varepsilon_{i, 1}$ and $\Pr(E = 1 | X = i, D = 0) = \varepsilon_{i, 0}$).
- Usually our data doesn't tell us who's in which sub-population, which can lead to misleading results\dots
- We can simulate the effects of population stratification by simulating and analyzing with \texttt{TriLLIEM}.

## Simulating population stratification

- For the hybrid model, this means we'd have a case control study, where case and control counts are predetermined.
- To simulate population stratification efficiently we assume that disease status ($D$) and environmental exposure ($E$) are conditionally independent on one's sub-population ($X$), then if we have two sub-populations of proportions $\omega_i$...
\begin{align*}\label{eq:popstratex}
  &\Pr(X = i, E = 1 \mid D = 1) \\
  = &\frac{\Pr(D = 1, E = 1 \mid X = i) \Pr(X = i)}{\Pr(D = 1)} \\
  = &\frac{\Pr(D = 1 \mid X = i) \Pr(E = 1 \mid X = i) \Pr(X = i)}{\Pr(D = 1)} \\
  = &\frac{\alpha_i \varepsilon_{i, 1} \omega_i}{\sum_{j = 1}^2 \alpha_j \omega_j }.
\end{align*}

## Simulating population stratification, cont'd

- We have the probabilities for each cell conditional on $D$, hence we can sample from two multinomial distributions with $4$ categories to generate counts for each cell.
- Then for each cell in the table, we can simulate the genotype data conditional on $E$, $D$, and $X$.
- Because of our assumption of conditional independence, when simulating population stratification, we \textbf{must} not have gene-environment interactions.

\begin{table}[]
\centering
\begin{tabular}{l|l|l|}
\cline{2-3}
                                     & $D = 1$ & $D = 0$ \\ \hline
\multicolumn{1}{|l|}{$E = 1, X = 1$} & $a$     & $e$     \\
\multicolumn{1}{|l|}{$E = 1, X = 2$} & $b$     & $f$     \\ \hline
\multicolumn{1}{|l|}{$E = 0, X = 1$} & $c$     & $g$     \\
\multicolumn{1}{|l|}{$E = 0, X = 2$} & $d$     & $h$     \\ \hline
\multicolumn{1}{|l|}{Total}          & $1000$  & $1000$  \\ \hline
\end{tabular}
\end{table}

## Simulation study parameters

- To evaluate the Type 1 error for \texttt{TriLLIEM} under population stratification, we simulated data without any gene effects.
- We ran the simulation $2000$ times for $1000$ case and control trios, using the demographic parameters in the table below.
    - For comparison, we also simulated data with identical parameters but with $2000$ cases only.
- Each simulated data set was analyzed with \texttt{TriLLIEM} under the HWE, MS, and MaS models (only HWE and MS for the case-only data), with $\textrm{C}$, $\textrm{M}$, and one of $\textrm{E}:\textrm{M}, E:I_{\rm M}, E:I_{\rm F}$ as the effects.

\begin{table}[htb!]
\centering
\begin{tabular}{llllll}
  \hline
  Parameter        & $\text{MAF}_{i}$   & $\varepsilon_{i, 1}$ & $\varepsilon_{i, 0}$ & $\alpha_i$ & $\omega_i$ \\ \hline
Sub-population 1 (i = 1)  & $0.2$ & $0.3$        & $0.3$        & $0.1$    & $0.4$    \\
Sub-population 2 (i = 2) & $0.4$ & $0.5$        & $0.5$        & $0.2$    & $0.6$    \\ \hline
\end{tabular}
\end{table}

## Simulation study results, with controls

```{r, include = FALSE}
source("../utils/get_bias.R")

conditions <- expand.grid(
  R = 1,
  S = 1,
  V = 1,
  mtCoef = 1,
  Im = 1,
  If = 1,
  ntrios = 2000,
  includeE = TRUE,
  Einteraction = c("M", "Im", "If"),
  includeControl = c(TRUE, FALSE),
  nControl = 1000,
  stringsAsFactors = FALSE
)
trill_PS_results <- readRDS("../data/TriLLIEM_res_PS.RDS")

bias_table <- get_trill_table(trill_PS_results, conditions)

get_reject <- function(results){
  if(is.null(results)) return(c("C" = NA, "M" = NA, "E:M" = NA, "E:Im" = NA, "E:If" = NA, "Im" = NA, "If" = NA))

  pvals <-
    mapply(
      FUN = function(mat){
        mat[,2]
      },
      mat = results
    ) %>%
    t() %>%
    apply(
      MARGIN = 2,
      FUN = function(x) (x < 0.05) %>% mean()
    )

  res <- c("C" = NA, "M" = NA, "E:M" = NA, "E:Im" = NA, "E:If" = NA, "Im" = NA, "If" = NA)
  elements <- intersect(names(pvals), c("C", "M", "E:M", "E:Im", "E:If", "Im", "If"))
  res[elements] <- pvals[elements]

  return(res)
}

mt <- c("HWE", "MS", "MaS")
reject_table <- lapply(
  X = 1:3,
  FUN = function(b){
    lapply(
      X = 1:nrow(conditions),
      FUN = function(a, b){
        if (is.null(trill_PS_results[[b]][[a]]$strat)) return(NULL)
        with(
          conditions[a,],
          rbind(get_reject(trill_PS_results[[b]][[a]]$strat), get_reject(trill_PS_results[[b]][[a]]$nostrat)) %>%
            as.data.frame() %>%
            mutate(strat = c("strat", "nostrat"),
                   includeControl = includeControl,
                    includeE = includeE,
                    mtCoef = mtCoef,
                    prE = if (exists("prE")) prE else NULL,
                    T.C = R,
                    T.M = S,
                    T.V = V,
                    T.Im = conditions[a, "Im"],
                    T.If = conditions[a, "If"]
                   )
        )
      },
      b = b
    ) %>%
      bind_rows() %>%
      mutate(mt = mt[[b]])
  }
) %>%
  bind_rows()

control_reject <- kbl(reject_table %>% filter(strat == "nostrat", includeControl) %>% select(C, M, "E:M", "E:Im", "E:If", Im, If, mt),
    format = "latex",
    align = "l",
    col.names = c("$\\textrm{C}$", "$\\textrm{M}$", "$\\textrm{E}:\\textrm{M}$", "$\\textrm{E}:I_{\\rm M}$", "$\\textrm{E}:I_{\\rm F}$", "$I_{\\rm M}$", "$I_{\\rm F}$", "Mating type"),
    digits = 5,
    label = "reject_tables_nonstrat",
    escape = FALSE,
    booktabs = TRUE,
    linesep = c("", "", "\\addlinespace"),
    position = "H") %>%
    kable_styling(latex_options="scale_down", position = "center")

control_bias <- kbl(bias_table %>% filter(strat == "nostrat", includeControl) %>% select(C, M, "E:M", "E:Im", "E:If", Im, If, mt),
    format = "latex",
    align = "l",
    col.names = c("$\\textrm{C}$", "$\\textrm{M}$", "$\\textrm{E}:\\textrm{M}$", "$\\textrm{E}:I_{\\rm M}$", "$\\textrm{E}:I_{\\rm F}$", "$I_{\\rm M}$", "$I_{\\rm F}$", "Mating type"),
    digits = 5,
    label = "bias_tables_nonstrat",
    escape = FALSE,
    booktabs = TRUE,
    linesep = c("", "", "\\addlinespace"),
    position = "H") %>%
    kable_styling(latex_options="scale_down", position = "center")

nocontrol_reject <- kbl(reject_table %>% filter(strat == "nostrat", !includeControl) %>% select(C, M, "E:M", "E:Im", "E:If", Im, If, mt),
    format = "latex",
    align = "l",
    col.names = c("$\\textrm{C}$", "$\\textrm{M}$", "$\\textrm{E}:\\textrm{M}$", "$\\textrm{E}:I_{\\rm M}$", "$\\textrm{E}:I_{\\rm F}$", "$I_{\\rm M}$", "$I_{\\rm F}$", "Mating type"),
    digits = 5,
    label = "reject_tables_strat",
    escape = FALSE,
    booktabs = TRUE,
    linesep = c("", "", "\\addlinespace"),
    position = "H") %>%
  kable_styling(latex_options="scale_down")

nocontrol_bias <- kbl(bias_table %>% filter(strat == "nostrat", !includeControl) %>% select(C, M, "E:M", "E:Im", "E:If", Im, If, mt),
    align = "l",
    col.names = c("$\\textrm{C}$", "$\\textrm{M}$", "$\\textrm{E}:\\textrm{M}$", "$\\textrm{E}:I_{\\rm M}$", "$\\textrm{E}:I_{\\rm F}$", "$I_{\\rm M}$", "$I_{\\rm F}$", "Mating type"),
    digits = 5,
    label = "bias_tables_strat",
    escape = FALSE,
    booktabs = TRUE,
    linesep = c("", "", "\\addlinespace"),
    position = "H") %>%
    kable_styling(latex_options="scale_down")
```

\begin{columns}
\begin{column}{0.7\textwidth}
`r control_reject`
`r control_bias`
\end{column}
\begin{column}{0.25\textwidth}
Top: False rejection rates at $\alpha = 0.05$.\\
\vspace{0.5cm}
Bottom: Average relative bias of effect.
\end{column}
\end{columns}

## Simulation study results, without controls

\begin{columns}
\begin{column}{0.7\textwidth}
`r nocontrol_reject`
`r nocontrol_bias`
\end{column}
\begin{column}{0.25\textwidth}
Top: False rejection rates at $\alpha = 0.05$.\\
\vspace{0.5cm}
Bottom: Average relative bias of effect.
\end{column}
\end{columns}

## Comparison to data without population stratification, with controls

```{r, echo = FALSE}
conditions <- expand.grid(
  maf = 0.3,
  R = c(1, 1.15),
  S = 1,
  V = c(1, 1.5, 1.6),
  mtCoef = c(0.85, 1, 1.15),
  Im = c(1, 1.2, 1.4),
  If = c(1, 1.2, 1.4),
  includeE = c(FALSE, TRUE),
  Einteraction = c("Im", "If"),
  prE = c(0.3, 0.5),
  includeControl = c(FALSE, TRUE),
  prControl = 0.5,
  stringsAsFactors = FALSE
) %>%
  filter(
    !(!includeE & !(Einteraction == "Im" & V == 1 & prE == 0.3)),
    (R == 1) + (S == 1) + (V == 1) + (Im == 1) + (If == 1) >= 4,
    !(If != 1 & (Im != 1 | Einteraction == "Im")),
    !(Im != 1 & Einteraction == "If")
  )
trill_results <- readRDS("../data/TriLLIEM_res.RDS")

trill_bias_table <- 
  trill_results %>% 
  get_trill_table(conditions) %>% 
  filter(T.C == 1, T.M == 1, T.V == 1, T.Im == 1, T.If == 1, includeE, strat == "nostrat", mtCoef == 1, prE == 0.3)

trill_reject_table <- lapply(
  X = 1:3,
  FUN = function(b){
    lapply(
      X = 1:nrow(conditions),
      FUN = function(a, b){
        if (is.null(trill_results[[b]][[a]]$strat)) return(NULL)
        with(
          conditions[a,],
          rbind(get_reject(trill_results[[b]][[a]]$strat), get_reject(trill_results[[b]][[a]]$nostrat)) %>%
            as.data.frame() %>%
            mutate(strat = c("strat", "nostrat"),
                   includeControl = includeControl,
                    includeE = includeE,
                    mtCoef = mtCoef,,
                    prE = if (exists("prE")) prE else NULL,
                    T.C = R,
                    T.M = S,
                    T.V = V,
                    T.Im = conditions[a, "Im"],
                    T.If = conditions[a, "If"]
                   )
        )
      },
      b = b
    ) %>%
      bind_rows() %>%
      mutate(mt = mt[[b]])
  }
) %>%
  bind_rows() %>%
  filter(T.C == 1, T.M == 1, T.V == 1, T.Im == 1, T.If == 1, includeE, strat == "nostrat", mtCoef == 1, prE == 0.3)

control_reject <- kbl(trill_reject_table %>% filter(includeControl) %>% select(C, M, "E:M", "E:Im", "E:If", Im, If, mt),
    format = "latex",
    align = "l",
    col.names = c("$\\textrm{C}$", "$\\textrm{M}$", "$\\textrm{E}:\\textrm{M}$", "$\\textrm{E}:I_{\\rm M}$", "$\\textrm{E}:I_{\\rm F}$", "$I_{\\rm M}$", "$I_{\\rm F}$", "Mating type"),
    digits = 5,
    label = "reject_tables_nonstrat",
    escape = FALSE,
    booktabs = TRUE,
    linesep = c("", "\\addlinespace"),
    position = "H") %>%
    kable_styling(latex_options="scale_down", position = "center")

control_bias <- kbl(trill_bias_table %>% filter(includeControl) %>% select(C, M, "E:M", "E:Im", "E:If", Im, If, mt),
    format = "latex",
    align = "l",
    col.names = c("$\\textrm{C}$", "$\\textrm{M}$", "$\\textrm{E}:\\textrm{M}$", "$\\textrm{E}:I_{\\rm M}$", "$\\textrm{E}:I_{\\rm F}$", "$I_{\\rm M}$", "$I_{\\rm F}$", "Mating type"),
    digits = 5,
    label = "bias_tables_nonstrat",
    escape = FALSE,
    booktabs = TRUE,
    linesep = c("", "\\addlinespace"),
    position = "H") %>%
    kable_styling(latex_options="scale_down", position = "center")

nocontrol_reject <- kbl(trill_reject_table %>% filter(!includeControl) %>% select(C, M, "E:M", "E:Im", "E:If", Im, If, mt),
    format = "latex",
    align = "l",
    col.names = c("$\\textrm{C}$", "$\\textrm{M}$", "$\\textrm{E}:\\textrm{M}$", "$\\textrm{E}:I_{\\rm M}$", "$\\textrm{E}:I_{\\rm F}$", "$I_{\\rm M}$", "$I_{\\rm F}$", "Mating type"),
    digits = 5,
    label = "reject_tables_strat",
    escape = FALSE,
    booktabs = TRUE,
    linesep = c("", "\\addlinespace"),
    position = "H") %>%
  kable_styling(latex_options="scale_down")

nocontrol_bias <- kbl(trill_bias_table %>% filter(!includeControl) %>% select(C, M, "E:M", "E:Im", "E:If", Im, If, mt),
    align = "l",
    col.names = c("$\\textrm{C}$", "$\\textrm{M}$", "$\\textrm{E}:\\textrm{M}$", "$\\textrm{E}:I_{\\rm M}$", "$\\textrm{E}:I_{\\rm F}$", "$I_{\\rm M}$", "$I_{\\rm F}$", "Mating type"),
    digits = 5,
    label = "bias_tables_strat",
    escape = FALSE,
    booktabs = TRUE,
    linesep = c("", "\\addlinespace"),
    position = "H") %>%
    kable_styling(latex_options="scale_down")
```
\begin{columns}
\begin{column}{0.7\textwidth}
`r control_reject`
`r control_bias`
\end{column}
\begin{column}{0.25\textwidth}
Top: False rejection rates at $\alpha = 0.05$.\\
\vspace{0.5cm}
Bottom: Average relative bias of effect.
\end{column}
\end{columns}

## Comparison to data without population stratification, without controls

\begin{columns}
\begin{column}{0.7\textwidth}
`r nocontrol_reject`
`r nocontrol_bias`
\end{column}
\begin{column}{0.25\textwidth}
Top: False rejection rates at $\alpha = 0.05$.\\
\vspace{0.5cm}
Bottom: Average relative bias of effect.
\end{column}
\end{columns}

## Conclusion

- Hybrid model does indeed suffer under population stratification whereas case-only model performs fine.
    - Despite this, Type 1 error rates are low for environmental interaction terms.
- These errors occur regardless of mating type model when controls are present.
- Consequences for analyzing populations where mating asymmetry is suspected.
