# Parallelization
- If we want to run a simple function many times, it's not great time-wise...
```{r}
## system.time returns how long a piece of code takes to run in seconds
dat <- rep(NA, 1000)
system.time({
  for (i in 1:1000) {
    dat[i] <- mean(runif(1000000, min = 0, max = 1))
  }
})
```
- When each of these function calls are independent, we can speed it up by using parallelization.
- Instead of having the 1,000 function calls be run one at a time, we can enlist $n$ of our CPU \emph{cores} to each run the function at the same time, making our code $n$ times faster.
- We can do this with the `parallel` package.  For paralellization in R, we have two methods:
  - FORK: the "better" method, works ONLY on UNIX based systems (Linux, Mac, NOT Windows).  Windows users can use WSL to get these UNIX features though, but need to install R for Linux.
  - PSOCK: the "worse" method, works for all systems.
- Now if we want to parallelize our for loop, we use the `foreach` and `doParallel` packages:

```{r}
## The basic incantation for FORK paralleling a for loop
## Loading necessary packages
# install.packages(c("foreach", "doParallel"))
library(foreach) # Provides the foreach loop which supports parallelizing
library(doParallel) # Provides the %dopar% operator for foreach
library(parallel)

detectCores() # Optional, checks how many cores your CPU has, this is the upper bound for number of parallels
n.cores <- 5 ## REPLACE WITH APPROPRIATE NUMBER OF CORES
cl <- makeCluster(n.cores, type = "FORK")
registerDoParallel(cl = cl)

system.time({
  dat_fork <- foreach(i = 1:1000, .combine = "c") %dopar% {
    mean(runif(1000000, min = 0, max = 1)) ## REPLACE WITH CODE TO PARALLELIZE
  }
})

stopCluster(cl = cl)
```

```{r}
## The basic incantation for PSOCK paralleling a for loop
## Loading necessary packages
# install.packages(c("foreach", "doParallel"))
library(foreach) # Provides the foreach loop which supports parallelizing
library(doParallel) # Provides the %dopar% operator for foreach
library(parallel)

detectCores() # Optional, checks how many cores your CPU has, this is the upper bound for number of parallels
n.cores <- 5 ## REPLACE WITH APPROPRIATE NUMBER OF CORES
cl <- makeCluster(n.cores, type = "PSOCK")
registerDoParallel(cl = cl)

system.time({
  dat_psock <- foreach(i = 1:1000, .combine = "c") %dopar% {
    mean(runif(1000000, min = 0, max = 1)) ## REPLACE WITH CODE TO PARALLELIZE
  }
})

stopCluster(cl = cl)
```
- Generally for this type of parallelization, we should use apply functions instead of for loops.
```{r}
# Non-parallelized *apply
# apply for matrices
# lapply lists
# sapply for vectors and lists with automatic simplification
# vapply for vectors and lists where you specify expected output
# mapply for multiple vectors/lists of inputs
system.time({
  dat_app <- mapply(FUN = function(x) mean(runif(1000000, min = 0, max = 1)),
                    x = 1:1000 # Arguments for FUN should go here
                    )
})
```
- We can easily parallelize apply functions, only requiring the `parallel` package:
```{r}
## The basic incantation for FORK parallelizing a *apply function
## With FORK, we get parallel lapply and mapply (mclapply and mcmapply)
## Load parallel package
library(parallel)

detectCores() # Optional, checks how many cores your CPU has, this is the upper bound for number of parallels
n.cores <- 10 ## REPLACE WITH APPROPRIATE NUMBER OF CORES

system.time({
  dat_app_fork <- mcmapply(
    FUN = function(x) mean(runif(1000000, min = 0, max = 1)),
    mc.cores = n.cores, ## If left blank, will automatically use all the cores available
    x = 1:1000
  )
})
```
- This also works for PSOCK, with a bit more set up:
```{r}
## The basic incantation for PSOCK parallelizing a *apply function
## With PSOCK, we get parallel apply, sapply, and lapply (parApply, parSapply, parLapply)
## Load parallel package
library(parallel)

detectCores() # Optional, checks how many cores your CPU has, this is the upper bound for number of parallels
n.cores <- 10 ## REPLACE WITH APPROPRIATE NUMBER OF CORES

cl <- makeCluster(n.cores, type = "PSOCK")

system.time({
  dat_app_psock <- parSapply(
    cl = cl,
    X = 1:1000,
    FUN = function(x) mean(runif(1000000, min = 0, max = 1))
  )
})

stopCluster(cl = cl)
```
