```{r setup, include = FALSE}
library(tidyverse); theme_set(theme_bw())
library(TriLLIEM)
library(parallel)
```

First, will the parameterization be the Haplin parameterization for PoO (i.e. using $C_m, C_f$)? If not then we can't include Haplin in the comparison.  Should we use parallelization in this script to speed things up?  We also need the path to EMIM as well.

```{r}
includeHaplin = TRUE # TRUE for Haplin, FALSE for no Haplin
includeParallel = TRUE # TRUE for parallelization, FALSE for no parallelization
cores <- 10 # Number of cores to use if includeParallel is TRUE
EMIM_path <- "./EMIM/" ## CHANGE THIS TO THE PATH FOR THE EMIM DIRECTORY
```

What type of data should we simulate?

```{r}
## Parameters we're focusing on now
R <- c(1,1,1) # Child effects, should be all 1's if including Haplin
S <- c(1,1,1) # Maternal effects
Im <- 1 # Maternal imprinting parameter/Cm parameter for Haplin
If <- 1 # Paternal imprinting parameter/Cf parameter for Haplin

## Parameters that could be useful to look at
mtmodel <- "HWE" # Hardy-Weinberg = HWE, Mating Symmetry = MS, Mating Asymmetry = MaS
maf <- 0.3 # Minor allele frequency
nsim <- 100 # Number of iid data sets to be generated
ntrios <- 1000 # Number of trios to include in each data set

## Parameters we'll be looking at later
includeE <- FALSE # TRUE for gene-environment interactions
V <- c(1,2,4) # Maternal gene-environment interactions effects
includeControl <- FALSE # TRUE for half controls, FALSE for no controls, not tested...
```

What effects are we including in the TriLLIEM/EMIM/Haplin models?

```{r}
## effects can include "C", "M", "Im", "If", but not all three of "C", "Im" and "If"
## Haplin parameterization REQUIRES "Im" and "If"
effects <- c("M", "Im", "If")
```

```{r analysis_functions, include = FALSE}
TRILL_analysis <- function(dat, effects){
  if (includeParallel){
    my.cluster <- makeCluster(cores)
    clusterEvalQ(my.cluster, {
      library(TriLLIEM)
      library(tidyverse)
    })
    
    clusterExport(my.cluster, c("effects", "mtmodel"))
    
    startT <- Sys.time()
    
    TriLL_res <- parLapply(
      my.cluster,
      dat,
      function(x){
        TriLLIEM(mtmodel = mtmodel,
                 dat = x$dat4R,
                 effects = effects,
                 max.iter = 30) %>%
          summ_trill(effects = effects) %>%
          unlist()
      }
    ) %>%
      do.call(rbind, .)
    
    endT <- Sys.time()
    print("trill:")
    print(endT - startT)
    stopCluster(cl = my.cluster)
  } else {
    TriLL_res <- lapply(
      dat,
      function(x){
        TriLLIEM(mtmodel = mtmodel,
                 dat = x$dat4R,
                 effects = effects,
                 max.iter = 30) %>%
          summ_trill(effects = effects) %>%
          unlist()
      }
    ) %>%
      do.call(rbind, .)
  }
  return(TriLL_res)
}

EMIM_analysis <- function(dat, effects){  
  if(includeParallel){
    my.cluster <- makeCluster(cores)
    clusterEvalQ(my.cluster, {
      library(TriLLIEM)
      library(tidyverse)
    })
      
    clusterExport(my.cluster, c("effects", "mtmodel", "EMIM_path"))
      
    startT <- Sys.time()
    
    emim_res <- parLapply(
      my.cluster,
      dat,
      function(x){
        runEMIM(mtmodel = mtmodel,
                peddat = x$dat4EMIM, 
                effects = effects,
                emimpath = EMIM_path) %>% 
          summ_emim() %>%
          unlist()
      }
    ) %>%
      do.call(rbind, .)
    
    endT <- Sys.time()
    print("emim:")
    print(endT - startT)
    stopCluster(cl = my.cluster)
  } else {
    emim_res <- lapply(
      dat,
      function(x){
        runEMIM(mtmodel = mtmodel,
                peddat = x$dat4EMIM, 
                effects = effects,
                emimpath = EMIM_path) %>% 
          summ_emim() %>%
          unlist()
      }
    ) %>%
      do.call(rbind, .)
  }
  return(emim_res)
}

Haplin_analyses <- function(dat, effects){
  if(includeParallel){
    my.cluster <- makeCluster(cores)
    clusterEvalQ(my.cluster, {
      library(TriLLIEM)
      library(tidyverse)
    })
    
    clusterExport(my.cluster, "effects")
      
    startT <- Sys.time()
    
    hap_res <- parLapply(
      my.cluster,
      dat,
      function(x){
        runHaplin(effects = effects,
                  dat = x$dat4haplin, 
                  PoO = TRUE) %>% 
          summ_haplin(PoO = TRUE) %>%
          unlist()
      }
    ) %>%
      do.call(rbind, .)
    
    endT <- Sys.time()
    print("haplin:")
    print(endT - startT)
    stopCluster(cl = my.cluster)
  } else {
    hap_res <- lapply(
      dat,
      function(x){
        runHaplin(effects = effects,
                  dat = x$dat4haplin, 
                  PoO = TRUE) %>% 
          summ_haplin(PoO = TRUE) %>%
          unlist()
      }
    ) %>%
      do.call(rbind, .)
  }
  return(hap_res)
}
```

Simulate the data:

```{r}
startT <- Sys.time()
if (includeParallel) {
  my.cluster <- makeCluster(cores)
    clusterEvalQ(my.cluster, {
      library(TriLLIEM)
      library(tidyverse)
    })
    
  clusterExport(my.cluster, 
                c("ntrios",
                  "nsim",
                  "R",
                  "S",
                  "Im",
                  "If",
                  "mtmodel",
                  "maf",
                  "includeE",
                  "V",
                  "includeControl")
                )
    
  dat <- parLapply(
    my.cluster,
    rep(ntrios, nsim),
    function(x){
      simulateData(
        ntrios = x, 
        R = R,
        S = S,
        Im = Im,
        If = If,
        mtmodel = mtmodel,
        maf = maf,
        includeE = includeE,
        V = V,
        includeControl = includeControl)
    }
  )
  stopCluster(cl = my.cluster)
} else {
  dat <- lapply(
    rep(ntrios, nsim),
    function(x){
      simulateData(
        ntrios = x, 
        R = R,
        S = S,
        Im = Im,
        If = If,
        mtmodel = mtmodel,
        maf = maf,
        includeE = includeE,
        V = V,
        includeControl = includeControl)
    }
  )
}
endT <- Sys.time()
print("data creation:")
print(endT - startT)
```

Analyze the data with TriLLIEM, EMIM, and Haplin (if Haplin's included):

```{r}
trill_res <- TRILL_analysis(dat = dat, effects = effects)
emim_res <- EMIM_analysis(dat = dat, effects = effects)
if(includeHaplin){
  hap_res <- Haplin_analyses(dat = dat, effects = effects)
}
```

Haplin does not seem to report standard error, only a p-value, but EMIM does return a standard error.

p-values for EMIM are calculated ourselves using the formula:

\[2 \cdot (1- \Phi(|\ln R_1 / {\rm SE}(\ln R_1)|) )\]

or as example code,
```{r, eval = FALSE}
2 * pnorm(abs(res$lnR1 / res$sd_lnR1), lower = F)
```
