---
title: "TriLLIEM"
subtitle: "Triad Log-Linear modelling of Imprinting, Environmental interactions, and Maternal effects"
author: Kevin Zhao
output:
  beamer_presentation:
    theme: "AnnArbor"
    colortheme: "dolphin"
    fonttheme: "structurebold"
header-includes: 
  - \usepackage{tikz}
  - \usepackage{pgfplots}
  - \usepackage{xcolor}
  - \pgfplotsset{compat=1.18}
  - \usepackage{mathptmx}
  - \usepackage[usenames,dvipsnames]{xcolor}
  - \let\oldShaded\Shaded
  - \let\endoldShaded\endShaded
  - \renewenvironment{Shaded}{\footnotesize\oldShaded}{\endoldShaded}
  - \let\oldverbatim\verbatim
  - \let\endoldverbatim\endverbatim
  - \renewenvironment{verbatim}{\footnotesize\oldverbatim}{\endoldverbatim}
---

```{r setup, include = FALSE}
library(tidyverse); theme_set(theme_bw())
library(TriLLIEM)
library(parallel)
```

## Triad data

- Goal: Quantifying the effect of a single-nucleotide polymorphism (SNP) of interest on one's disease risk.
- Case-control studies (\color{blue}Control \color{black}+ \color{red}Case\color{black}) are decent, but influenced by population substructures.
- Case-parent/Triad studies (\color{blue}Control Parents \color{black}+ \color{red}Case Child\color{black}) address this, and can identify \emph{maternal effects}.  Can't quantify mating asymmetry with case-triads alone.
- A \textbf{hybrid} triad study with \color{blue}control \color{black}and \color{red}case \color{black}children can solve this issue.

## The hybrid model

- The counts of each triad are modelled with a log-linear GLM.
- A simple example of a model we can apply to data:
\[\ln(n_{M,F,C,D}) = \ln (\mu_i) + CD \ln R + MD \ln S + \ln (\mathrm{Off}).\]
- Additional terms to consider:
    - Mating type parameters (for HWE, MS, and MaS)
    - Imprinting effects ($I_m$ and/or $I_f$)
    - Gene-environment interactions ($M:E$, $I_m:E$, $I_f:E$, $C:E$)

## TriLLIEM in action!

```{r}
dat <- simulateData(
  ntrios = 10000, # Simulate 10,000 triads
  mtmodel = "MaS", # Simulate Mating Asymmetry data
  mtCoef = c(1.15, 1.15, 1), # MaS parameters
  R = c(2^0, 2^1, 2^2), # Risk if child has 0, 1, or 2 copies
  S = c(1.5^0, 1.5^1, 1.5^2), # Risk if mother has 0, 1, or 2 copies
  V = c(1.1^0, 1.1^1, 1.1^2), # Maternal gene-environment risk
  Im = 2.5, # Maternal imprinting increases risk 2.5-fold
  includeE = TRUE, # Including environmental effects
  prE = 0.3, # 30% of the sample will have E = 1
  includeControl = TRUE, # Including controls
  prControl = 0.3 # 30% of the sample will have D = 0
)$dat4R
```
## TriLLIEM in action!

```{r}
head(dat, n = 16)
```

## TriLLIEM in action!

```{r}
results <- TriLLIEM(
  mtmodel = "MaS", 
  dat = dat, 
  effects = c("C", "M", "Im"),
  includeE = TRUE,
  includeD = TRUE
) %>% 
  summary() %>%
  coef() %>%
  .[c("C", "M", "Im", "E:M"), c(1,4)]
results[, 1] <- exp(results[, 1])
results
```

## TriLLIEM in action!

```{r}
results <- TriLLIEM(
  mtmodel = "HWE", 
  dat = dat, 
  effects = c("C", "M", "Im"),
  includeE = TRUE,
  includeD = TRUE
) %>% 
  summary() %>%
  coef() %>%
  .[c("C", "M", "Im", "E:M"), c(1,4)]
results[, 1] <- exp(results[, 1])
results
```

## Verifying TriLLIEM works!

```{r datsim, cache=TRUE}
dat <- mclapply(
  X = 1:1000, # Simulate 1000 times
  FUN = function(X, ...) simulateData(...)$dat4R, # What we're repeating
  ntrios = 1000, # Simulate 10,000 triads
  mtmodel = "MaS", # Simulate Mating Asymmetry data
  mtCoef = c(1.15, 1.15, 1), # MaS parameters
  R = c(2^0, 2^1, 2^2), # Risk if child has 0, 1, or 2 copies
  S = c(1.5^0, 1.5^1, 1.5^2), # Risk if mother has 0, 1, or 2 copies
  V = c(1.1^0, 1.1^1, 1.1^2), # Maternal gene-environment risk
  Im = 2.5, # Maternal imprinting increases risk 2.5-fold
  includeE = TRUE, # Including environmental effects
  prE = 0.3, # 30% of the sample will have E = 1
  includeControl = TRUE, # Including controls
  prControl = 0.3, # 30% of the sample will have D = 0
  mc.cores = 10
)
```

## Verifying TriLLIEM works!

```{r masres, cache=TRUE}
results_MaS <- mclapply(X = dat, function(X) {
  results <- TriLLIEM(
    dat = X,
    mtmodel = "MaS",
    effects = c("C", "M", "Im"),
    includeE = TRUE,
    includeD = TRUE
  ) %>%
    summary() %>%
    coef() %>%
    .[c("C", "M", "Im", "E:M"), c(1, 4)]
  results[, 1] <- exp(results[, 1])
  return(results)
}, mc.cores = 10)
```

## Verifying TriLLIEM works!

```{r masplot, fig.height = 3, cache=TRUE}
sapply(X = results_MaS, FUN = function(X) X[,1]) %>% 
  t() %>% 
  boxplot(horizontal = TRUE, col = palette.colors(5)[-1], las = 1)
abline(v = 2, col = palette.colors(5)[2])
abline(v = 1.5, col = palette.colors(5)[3])
abline(v = 2.5, col = palette.colors(5)[4])
abline(v = 1.1, col = palette.colors(5)[5])
## How often do we properly claim a parameter is relevant?
sapply(X = results_MaS, FUN = function(X) X[,2]) %>%
  apply(MARGIN = 1, FUN = function(x) sum(x < 0.05)/length(x))
```

## Simulating violations of the HW assumption 

```{r hweres, cache=TRUE}
results_HWE <- mclapply(X = dat, function(X) {
  results <- TriLLIEM(
    dat = X,
    mtmodel = "HWE", # The only change!
    effects = c("C", "M", "Im"),
    includeE = TRUE,
    includeD = TRUE
  ) %>%
    summary() %>%
    coef() %>%
    .[c("C", "M", "Im", "E:M"), c(1, 4)]
  results[, 1] <- exp(results[, 1])
  return(results)
}, mc.cores = 10)
```

## Simulating violations of the HW assumption 

```{r hweplot, fig.height = 3, cache=TRUE}
sapply(X = results_HWE, FUN = function(X) X[,1]) %>% 
  t() %>% 
  boxplot(horizontal = TRUE, col = palette.colors(5)[-1], las = 1)
abline(v = 2, col = palette.colors(5)[2])
abline(v = 1.5, col = palette.colors(5)[3])
abline(v = 2.5, col = palette.colors(5)[4])
abline(v = 1.1, col = palette.colors(5)[5])
## How often do we properly claim the parameters are relevant?
sapply(X = results_HWE, FUN = function(X) X[,2]) %>%
  apply(MARGIN = 1, FUN = function(x) sum(x < 0.05)/length(x))
```

## Results with null data

```{r datnullsim, cache=TRUE}
dat_null <- mclapply(
  X = 1:1000, # Simulate 1000 times
  FUN = function(X, ...) simulateData(...)$dat4R, # What we're repeating
  ntrios = 1000, # Simulate 10,000 triads
  mtmodel = "MaS", # Simulate Mating Asymmetry data
  mtCoef = c(1.15, 1.15, 1), # MaS parameters
  includeE = TRUE, # Including environmental effects
  prE = 0.3, # 30% of the sample will have E = 1
  includeControl = TRUE, # Including controls
  prControl = 0.3, # 30% of the sample will have D = 0
  mc.cores = 10
)
```

## Results with null data

```{r masresnull, cache=TRUE}
results_MaS_null <- mclapply(X = dat_null, function(X) {
  results <- TriLLIEM(
    dat = X,
    mtmodel = "MaS",
    effects = c("C", "M", "Im"),
    includeE = TRUE,
    includeD = TRUE
  ) %>%
    summary() %>%
    coef() %>%
    .[c("C", "M", "Im", "E:M"), c(1, 4)]
  results[, 1] <- exp(results[, 1])
  return(results)
}, mc.cores = 10)
```

## Results with null data

```{r masplotnull, fig.height = 3, cache=TRUE}
sapply(X = results_MaS_null, FUN = function(X) X[,1]) %>% 
  t() %>% 
  boxplot(horizontal = TRUE, col = palette.colors(5)[-1], las = 1)
## How often do we erroneously claim a parameter is significant?
sapply(X = results_MaS_null, FUN = function(X) X[,2]) %>%
  apply(MARGIN = 1, FUN = function(x) sum(x < 0.05)/length(x))
```

## Results with null data

```{r hweresnull, cache=TRUE}
results_HWE_null <- mclapply(X = dat_null, function(X) {
  results <- TriLLIEM(
    dat = X,
    mtmodel = "HWE", # The only change!
    effects = c("C", "M", "Im"),
    includeE = TRUE,
    includeD = TRUE
  ) %>%
    summary() %>%
    coef() %>%
    .[c("C", "M", "Im", "E:M"), c(1, 4)]
  results[, 1] <- exp(results[, 1])
  return(results)
}, mc.cores = 10)
```

## Results with null data

```{r hweplotnull, fig.height = 3, cache=TRUE}
sapply(X = results_HWE_null, FUN = function(X) X[,1]) %>% 
  t() %>% 
  boxplot(horizontal = TRUE, col = palette.colors(5)[-1], las = 1)
## How often do we erroneously claim a parameter is significant?
sapply(X = results_HWE_null, FUN = function(X) X[,2]) %>%
  apply(MARGIN = 1, FUN = function(x) sum(x < 0.05)/length(x))
```
