---
title: "TriLLIEM"
subtitle: "Triad Log-Linear modelling of Imprinting, Environmental interactions, and Maternal effects"
author: Kevin Zhao
output:
  beamer_presentation:
    theme: "AnnArbor"
    colortheme: "dolphin"
    fonttheme: "structurebold"
header-includes: 
  - \usepackage{tikz}
  - \usepackage{pgfplots}
  - \usepackage{xcolor}
  - \pgfplotsset{compat=1.18}
  - \usepackage{mathptmx}
  - \usepackage[usenames,dvipsnames]{xcolor}
  - \let\oldShaded\Shaded
  - \let\endoldShaded\endShaded
  - \renewenvironment{Shaded}{\footnotesize\oldShaded}{\endoldShaded}
  - \let\oldverbatim\verbatim
  - \let\endoldverbatim\endverbatim
  - \renewenvironment{verbatim}{\footnotesize\oldverbatim}{\endoldverbatim}
---

```{r setup, include = FALSE}
library(tidyverse); theme_set(theme_bw())
```

## Triad data

- Goal: Quantifying the effect of a single-nucleotide polymorphism (SNP) of interest on one's disease risk.
- Case-control studies (\color{blue}Control \color{black}+ \color{red}Case\color{black}) are decent, but influenced by population substructures.
- Case-parent/Triad studies (\color{blue}Control Parents \color{black}+ \color{red}Case Child\color{black}) address this, and can identify \emph{maternal effects}.  Can't quantify mating asymmetry with case-triads alone.
- A \textbf{hybrid} triad study with \color{blue}control \color{black}and \color{red}case \color{black}children can solve these issues.

## TriLLIEM in action!

```{r}
library(TriLLIEM)
## R = child effects for 0, 1, 2 copies of minor allele
dat <- simulateData(R = c(2^0, 2^1, 2^2)) 
TriLLIEM(dat = dat$dat4R, effects = c("C", "M")) %>% 
  summ_trill(effects = c("C", "M"))
```

## The hybrid model

- The counts of each triad are modelled with a log-linear GLM.
- A simple example of a model we can apply to data:
\[\ln(n_{M,F,C}) = \ln K + C \ln R + M \ln S + \ln \Pr(C|M,F).\]
- Additional terms to consider:
    - Mating type parameters
    - Gene-environment interactions
    - Imprinting effects

## Comparisons with EMIM - imprinting data

- 2000 data sets, 1000 trios each with maternal, child, and imprinting effects.
- Simulated null model with $M = C = I_m = 1$, what do both methods give?

```{r, echo = FALSE, fig.height = 3}
trill_null <- data.frame(readRDS("data_EMIM/trill_NULL_CMI1.rds"))
emim_null <- data.frame(readRDS("data_EMIM/EMIM_NULL_CMI1.rds"))

ggplot(trill_null %>%
         mutate(EMIM.effects.C = emim_null$effects.C,
                EMIM.effects.M = emim_null$effects.M,
                EMIM.effects.Im = emim_null$effects.Im) %>%
         pivot_longer(c(effects.C,
                        EMIM.effects.C,
                        effects.M,
                        EMIM.effects.M,
                        effects.Im,
                        EMIM.effects.Im),
                      names_to = "Parameter",
                      values_to = "Estimate") %>%
         mutate(Parameter = fct_relevel(Parameter,
                                        rev(
                                          c(
                                            "effects.C",
                                            "EMIM.effects.C",
                                            "effects.M",
                                            "EMIM.effects.M",
                                            "effects.Im",
                                            "EMIM.effects.Im"
                                          )
                                        )))
) +
  geom_boxplot(aes(x = Estimate, y = Parameter)) +
  scale_y_discrete(
    labels = rev(c("TriLLIEM C",
               "EMIM C",
               "TriLLIEM M",
               "EMIM M",
               "TriLLIEM Im",
               "EMIM Im"))
  ) +
  ylab("")
## Significances should be sim, check SE.  also check diff of estimates between im
## EMIM uses a direct maximization approach for splitting (1,1,1) cell, and for dyads.
## Check Ainsworth, used direct search algorithm
```
```{r, include = FALSE}
sum(trill_null$pvals.Im < 0.05)/nrow(trill_null)
sum(emim_null$pvals.Im < 0.05)/nrow(emim_null)
```
- At this sample size, TriLLIEM incorrectly deemed $I_m$ significant at $\alpha = 0.05$ 10.8\% of the time, while EMIM deemed it significant 5.6\% of the time.

## Comparisons with EMIM

- 2000 data sets, 1000 trios each with child, maternal, and imprinting effects ($C = M = I_m = 2$).

```{r, echo = FALSE, fig.height = 3}
trill_alt1 <- data.frame(readRDS("data_EMIM/trill_ALT_CMI2.rds"))
emim_alt1 <- data.frame(readRDS("data_EMIM/EMIM_ALT_CMI2.rds"))

ggplot(trill_alt1 %>%
         mutate(EMIM.effects.C = emim_alt1$effects.C,
                EMIM.effects.M = emim_alt1$effects.M,
                EMIM.effects.Im = emim_alt1$effects.Im) %>%
         pivot_longer(c(effects.C,
                        EMIM.effects.C,
                        effects.M,
                        EMIM.effects.M,
                        effects.Im,
                        EMIM.effects.Im),
                      names_to = "Parameter",
                      values_to = "Estimate") %>%
         mutate(Parameter = fct_relevel(Parameter,
                                        rev(
                                          c(
                                            "effects.C",
                                            "EMIM.effects.C",
                                            "effects.M",
                                            "EMIM.effects.M",
                                            "effects.Im",
                                            "EMIM.effects.Im"
                                          )
                                        )))
) +
  geom_boxplot(aes(x = Estimate, y = Parameter)) +
  scale_y_discrete(
    labels = rev(c("TriLLIEM C",
               "EMIM C",
               "TriLLIEM M",
               "EMIM M",
               "TriLLIEM Im",
               "EMIM Im"))
  ) +
  ylab("")
```
```{r, include = FALSE}
sum(trill_alt1$pvals.Im > 0.05)/nrow(trill_alt1)
sum(emim_alt1$pvals.Im > 0.05)/nrow(emim_alt1)
```
- At this sample size, TriLLIEM failed to deem $I_m$ significant at $\alpha = 0.05$ 0.85\% of the time, while EMIM failed to deem it significant 37.8\% of the time.

## Comparisons with EMIM

- 2000 data sets, 1000 trios each with maternal imprinting effects only ($C = M = 1, I_m = 2$).

```{r, echo = FALSE, fig.height = 3}
trill_alt2 <- data.frame(readRDS("data_EMIM/trill_ALT_CM1_I2.rds"))
emim_alt2 <- data.frame(readRDS("data_EMIM/EMIM_ALT_CM1_I2.rds"))

ggplot(trill_alt2 %>%
         mutate(EMIM.effects.C = emim_alt2$effects.C,
                EMIM.effects.M = emim_alt2$effects.M,
                EMIM.effects.Im = emim_alt2$effects.Im) %>%
         pivot_longer(c(effects.C,
                        EMIM.effects.C,
                        effects.M,
                        EMIM.effects.M,
                        effects.Im,
                        EMIM.effects.Im),
                      names_to = "Parameter",
                      values_to = "Estimate") %>%
         mutate(Parameter = fct_relevel(Parameter,
                                        rev(
                                          c(
                                            "effects.C",
                                            "EMIM.effects.C",
                                            "effects.M",
                                            "EMIM.effects.M",
                                            "effects.Im",
                                            "EMIM.effects.Im"
                                          )
                                        )))
) +
  geom_boxplot(aes(x = Estimate, y = Parameter)) +
  scale_y_discrete(
    labels = rev(c("TriLLIEM C",
               "EMIM C",
               "TriLLIEM M",
               "EMIM M",
               "TriLLIEM Im",
               "EMIM Im"))
  ) +
  ylab("")
```
```{r, include = FALSE}
sum(trill_alt2$pvals.Im > 0.05)/nrow(trill_alt2)
sum(emim_alt2$pvals.Im > 0.05)/nrow(emim_alt2)
```
- At this sample size, TriLLIEM failed to deem $I_m$ significant at $\alpha = 0.05$ 0.4\% of the time, while EMIM failed to deem it significant 9.4\% of the time.

## Comparisons with Haplin

- 2000 data sets, 1000 trios each with maternal, and imprinting effects.
- Simulated null model with $M = C_m = C_f = 1$, what do both methods give?

```{r, echo = FALSE, fig.height = 3}
trill_null <- data.frame(readRDS("data_Haplin/trill_NULL_MII1.rds"))
hap_null <- data.frame(readRDS("data_Haplin/Haplin_NULL_MII1.rds"))

ggplot(trill_null %>%
         mutate(Haplin.effects.M = hap_null$effects.M,
                Haplin.effects.Im = hap_null$effects.RRcm,
                Haplin.effects.If = hap_null$effects.RRcf) %>%
         pivot_longer(c(effects.M,
                        Haplin.effects.M,
                        effects.Im,
                        Haplin.effects.Im,
                        effects.If,
                        Haplin.effects.If),
                      names_to = "Parameter",
                      values_to = "Estimate") %>%
         mutate(Parameter = fct_relevel(Parameter,
                                        rev(
                                          c(
                                            "effects.M",
                                            "Haplin.effects.M",
                                            "effects.Im",
                                            "Haplin.effects.Im",
                                            "effects.If",
                                            "Haplin.effects.If"
                                          )
                                        )))
) +
  geom_boxplot(aes(x = Estimate, y = Parameter)) +
  scale_y_discrete(
    labels = rev(c("TriLLIEM M",
               "Haplin M",
               "TriLLIEM Cm",
               "Haplin Cm",
               "TriLLIEM Cf",
               "Haplin Cf"))
  ) +
  ylab("")
```

```{r, include = FALSE}
sum(trill_null$pvals.Im < 0.05)/nrow(trill_null)
sum(hap_null$pvals.RRcm < 0.05)/nrow(hap_null)
```
- At this sample size, TriLLIEM incorrectly deemed $C_m$ significant at $\alpha = 0.05$ 8.65\% of the time, while Haplin deemed it significant 4.9\% of the time.


## Comparisons with Haplin

- Simulated model with $M = C_m = C_f = 2$.

```{r, echo = FALSE, fig.height = 3}
trill_alt1 <- data.frame(readRDS("data_Haplin/trill_ALT_MII2.rds"))
hap_alt1 <- data.frame(readRDS("data_Haplin/Haplin_ALT_MII2.rds"))

ggplot(trill_alt1 %>%
         mutate(Haplin.effects.M = hap_alt1$effects.M,
                Haplin.effects.Im = hap_alt1$effects.RRcm,
                Haplin.effects.If = hap_alt1$effects.RRcf) %>%
         pivot_longer(c(effects.M,
                        Haplin.effects.M,
                        effects.Im,
                        Haplin.effects.Im,
                        effects.If,
                        Haplin.effects.If),
                      names_to = "Parameter",
                      values_to = "Estimate") %>%
         mutate(Parameter = fct_relevel(Parameter,
                                        rev(
                                          c(
                                            "effects.M",
                                            "Haplin.effects.M",
                                            "effects.Im",
                                            "Haplin.effects.Im",
                                            "effects.If",
                                            "Haplin.effects.If"
                                          )
                                        )))
) +
  geom_boxplot(aes(x = Estimate, y = Parameter)) +
  scale_y_discrete(
    labels = rev(c("TriLLIEM M",
               "Haplin M",
               "TriLLIEM Cm",
               "Haplin Cm",
               "TriLLIEM Cf",
               "Haplin Cf"))
  ) +
  ylab("")
```

```{r, include = FALSE}
sum(trill_alt1$pvals.Im > 0.05)/nrow(trill_alt1)
sum(hap_alt1$pvals.RRcm > 0.05)/nrow(hap_alt1)
```
- At this sample size, TriLLIEM and Haplin both had no false negatives for $C_m$.

## Comparisons with Haplin

- Simulated model with $M = C_f = 1$, $C_m = 2$.

```{r, echo = FALSE, fig.height = 3}
trill_alt2 <- data.frame(readRDS("data_Haplin/trill_ALT_MIf1_Im2.rds"))
hap_alt2 <- data.frame(readRDS("data_Haplin/Haplin_ALT_MIf1_Im2.rds"))

ggplot(trill_alt2 %>%
         mutate(Haplin.effects.M = hap_alt2$effects.M,
                Haplin.effects.Im = hap_alt2$effects.RRcm,
                Haplin.effects.If = hap_alt2$effects.RRcf) %>%
         pivot_longer(c(effects.M,
                        Haplin.effects.M,
                        effects.Im,
                        Haplin.effects.Im,
                        effects.If,
                        Haplin.effects.If),
                      names_to = "Parameter",
                      values_to = "Estimate") %>%
         mutate(Parameter = fct_relevel(Parameter,
                                        rev(
                                          c(
                                            "effects.M",
                                            "Haplin.effects.M",
                                            "effects.Im",
                                            "Haplin.effects.Im",
                                            "effects.If",
                                            "Haplin.effects.If"
                                          )
                                        )))
) +
  geom_boxplot(aes(x = Estimate, y = Parameter)) +
  scale_y_discrete(
    labels = rev(c("TriLLIEM M",
               "Haplin M",
               "TriLLIEM Cm",
               "Haplin Cm",
               "TriLLIEM Cf",
               "Haplin Cf"))
  ) +
  ylab("")
```

```{r, include = FALSE}
sum(trill_alt2$pvals.Im > 0.05)/nrow(trill_alt2)
sum(hap_alt2$pvals.RRcm > 0.05)/nrow(hap_alt2)
```
- At this sample size, TriLLIEM and Haplin both had no false negatives for $C_m$.